---
title: "Phylogenetic Signal in INLA"
author: "R package build"
date: '2022-07-15'
slug: []
categories: []
tags: []
Description: ''
Tags: []
Categories: []
DisableComments: no
---

# Phylogenetic signal in INLA

Recently, my research has been focused on quantifying the importance of processes that might drive the patterns of similarity we observe within a particular variable - the autocorrelation.

For the most part, I have been using [Gaussian Process models](https://www.youtube.com/watch?v=PIuqxOBJqLU), and relying on the excellent example provided by Paul Bürkner showing how to [estimate phylogenetic signal in BRMS](https://cran.r-project.org/web/packages/brms/vignettes/brms_phylogenetics.html).

For various reasons, I shifted to using a different software package, [INLA](https://www.r-inla.org/), and to ensure I understand the dynamics of this new system, I replicated Paul Bürkner's example in INLA with the advice and guidance of Russell Dinnage.

To remind us of the premise I copy the introduction below:

> ## Introduction

> In the present vignette, we want to discuss how to specify phylogenetic multilevel models using brms. These models are relevant in evolutionary biology when data of many species are analyzed at the same time. The usual approach would be to model species as a grouping factor in a multilevel model and estimate varying intercepts (and possibly also varying slopes) over species. However, species are not independent as they come from the same phylogenetic tree and we thus have to adjust our model to incorporate this dependency. The examples discussed here are from chapter 11 of the book Modern Phylogenetic Comparative Methods and the application in Evolutionary Biology (de Villemeruil & Nakagawa, 2014). The necessary data can be downloaded from the corresponding website (<https://www.mpcm-evolution.com/>). Some of these models may take a few minutes to fit.

> ### A Simple Phylogenetic Model

> Assume we have measurements of a phenotype, phen (say the body size), and a cofactor variable (say the temperature of the environment). We prepare the data using the following code.

We are going to use the same data as Paul, so we can compare outcomes:

```{r}
library(INLA)
library(ape)
library(ggplot2)

phylo <- ape::read.nexus("https://paul-buerkner.github.io/data/phylo.nex")
data_simple <- read.table(
  "https://paul-buerkner.github.io/data/data_simple.txt",
  header = TRUE
)
```

To model the relationships between societies, BRMS requires a covariance matrix, but INLA uses precision matrices. So, where BRMS uses `ape::vcv.phylo` to build a covariance matrix, we instead used `MCMCglmm::inverseA`.

```{r}
phylo_prec_mat = MCMCglmm::inverseA(phylo,
                                   nodes = "ALL",
                                   scale = TRUE)$Ainv
```

You may notice that the precision matrix we built has many more columns and rows than we have taxa. This is because of the setting `nodes = "ALL"`, which builds the precision matrix for the relationships between all taxa AND nodes. Building the precision matrix this way creates a sparse matrix, and makes the model estimation more computationally efficient.

However, this means that unlike BRMS, the rows of the phylogenetic matrix do not line up with the rows of the dataset we are going to model. So we need to match the rows in to the rows in `data_simple` to the rows in `phylo_prec_mat`.

```{r}
phylo_id <- match(phylo$tip.label, rownames(phylo_prec_mat))
```

Finally, we build the model using some defauly INLA settings.

```{r}
model_simple = inla(phen ~ cofactor +
                    f(phylo_id,
                      model = "generic0",
                      Cmatrix = phylo_prec_mat),
                  data = data_simple,
                  control.compute=list(config = TRUE))
```

INLA and BRMS have similar model construction. Both approaches use the standard model formula `y ~ x`, which is how the response and fixed co-factor effect are represented. INLA and BRMS differ in how they represent random effects. In BRMS the random effect is shown like so: `(1|gr(phylo, cov = A)`. The `1|` indicates varying intercepts, followed by the grouping function `gr` which acts as tool to insert and align the observations `phylo` with the covariance matrix `cov = A`. INLA contains a similar random effect set-up tool called `f` for function. `f` models random intercepts if only one categorical variable is given, and random slopes if a second continuous co-variate is given. `f`  is more versatile than the `gr` function in `brms`, and is used to model a wide range of possible Gaussian models, and therefore more arguments. The second argument within `f` is `model`, which has 50 possibilities and defines how the precision estimates are calculated (see `names(inla.models()$latent)` for all models). We use the [`generic0` model](https://inla.r-inla-download.org/r-inla.org/doc/latent/generic0.pdf), which I explain in more detail below. The final argument is `Cmatrix` which contains our phylogenetic precision matrix. 

First we show the INLA output

```{r}
summary(model_simple)
```

Followed by the BRMS output:

```
 Family: gaussian 
  Links: mu = identity; sigma = identity 
Formula: phen ~ cofactor + (1 | gr(phylo, cov = A)) 
   Data: data_simple (Number of observations: 200) 
  Draws: 4 chains, each with iter = 2000; warmup = 1000; thin = 1;
         total post-warmup draws = 4000

Group-Level Effects: 
~phylo (Number of levels: 200) 
              Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sd(Intercept)    14.57      2.14    10.45    18.90 1.00      873     1347

Population-Level Effects: 
          Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
Intercept    38.16      7.04    24.55    51.82 1.00     2286     2438
cofactor      5.18      0.14     4.91     5.45 1.00     6730     3471

Family Specific Parameters: 
      Estimate Est.Error l-95% CI u-95% CI Rhat Bulk_ESS Tail_ESS
sigma     9.19      0.72     7.85    10.68 1.00     1107     1898

Draws were sampled using sampling(NUTS). For each parameter, Bulk_ESS
and Tail_ESS are effective sample size measures, and Rhat is the potential
scale reduction factor on split chains (at convergence, Rhat = 1).
```

INLA prints two main areas of output we are interested in: `Fixed effects` (titled `Populations-Level Effects` in BRMS output) and `Model hyperparameters` (`Group-Level Effects`). Both INLA and BRMS estimate fixed effects in similar ways, and the results are comparable between these two approaches. 

The hyperparameters for random effects in INLA are precision estimates, as opposed to variance estimates in BRMS. So, they are not directly comparable. Fortunately, precision estimates are easily converted to the more intuitive variance estimates using the formula $\sqrt(1 / \tau)$. After this conversion, we see that INLA estimates a slightly smaller phylogenetic effect than BRMS does. 

```{r}
round(sqrt(1/model_simple$summary.hyperpar), 2)
```
The most intuitive aspect of the BRMS example is presenting the posterior distribution of phylogenetic signal. To produce the same plot from the INLA model, we can use the following code:

```{r}
posterior = inla.hyperpar.sample(n = 1000, model_simple)
lambda_estimates = data.frame(1 / posterior / rowSums(1 / posterior))

ggplot(lambda_estimates, aes(x = Precision.for.phylo_id)) + 
  geom_density(fill = "#6591b8") + 
  xlim(c(0.3, 0.9)) + # same limits as BRMS example
  ylab(element_blank()) + 
  theme_classic() + 
  theme(axis.text.y = element_blank(),
        axis.ticks = element_blank())
```

The INLA model performs quite well using default priors, with similar estimates, but slightly higher error. One of the more attractive aspects of INLA is the introduction of [penalizing complexity priors](https://becarioprecario.bitbucket.io/inla-gitbook/ch-priors.html#sec:pcpriors). The basic idea behind penalizing complexity priors is that we should prefer a simpler model, until there is enough support for a more complex model.
